---
title: "Neurotyping Analysis"
author: "Kenny Miller"
date: "5/5/2021"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
setwd("~/MSBA/UT AD Work/Neurotyping")
load("~/MSBA/UT AD Work/Neurotyping/neurotyping function.RData")
results <- read.csv("neurotyping_test.csv")
questions <- readxl::read_xlsx("Neurotyping Overall.xlsx")
results2 <- results[,-c(1:3)]
soccer_results <- read.csv("Neurotyping Test (Responses) - Form Responses 1.csv")
```

**This file includes the code and necessary functions to complete the neurotype analysis as requested by our team. All code is proprietary and not to be shared without the written consent of the author.**


## Matrix Assignment

```{r matrix creation, include=FALSE}
colnames(questions)[10:14] <- c("x5","x3","x1","x3","x5")
colnames(results2)[1]
q1 <- as.matrix(questions[2:6,10:14]); colnames(q1) <- 1:5; q1
colnames(results2)[2]
q2 <- as.matrix(questions[2:6,10:14]); colnames(q2) <- 1:5; q2
colnames(results2)[3]
q3 <- as.matrix(questions[7:11,10:14]); colnames(q3) <- 1:5; q3
colnames(results2)[4]
q4 <- as.matrix(questions[12:16,10:14]); colnames(q4) <- 1:5; q4
colnames(results2)[5]
q5 <- as.matrix(questions[17:11,10:14]); colnames(q5) <- 1:5; q5
colnames(results2)[6]
q6 <- as.matrix(questions[22:26,10:14]); colnames(q6) <- 1:5; q6
colnames(results2)[7]
q7 <- as.matrix(questions[22:26,10:14]); colnames(q7) <- 1:5; q7
colnames(results2)[8]
q8 <- as.matrix(questions[27:31,10:14]); colnames(q8) <- 1:5; q8
colnames(results2)[9]
q9 <- as.matrix(questions[27:31,10:14]); colnames(q9) <- 1:5; q9
colnames(results2)[10]
q10 <- as.matrix(questions[32:36,10:14]); colnames(q10) <- 1:5; q10
colnames(results2)[11]
q11 <- as.matrix(questions[37:41,10:14]); colnames(q11) <- 1:5; q11
colnames(results2)[12]
q12 <- as.matrix(questions[42:46,10:14]); colnames(q12) <- 1:5; q12
colnames(results2)[13]
q13 <- as.matrix(questions[42:46,10:14]); colnames(q13) <- 1:5; q13
colnames(results2)[14]
q14 <- as.matrix(questions[47:51,10:14]); colnames(q14) <- 1:5; q14
colnames(results2)[15]
q15 <- as.matrix(questions[47:51,10:14]); colnames(q15) <- 1:5; q15
colnames(results2)[16]
q16 <- as.matrix(questions[52:56,10:14]); colnames(q16) <- 1:5; q16
colnames(results2)[17]
q17 <- as.matrix(questions[57:61,10:14]); colnames(q17) <- 1:5; q17
colnames(results2)[18]
q18 <- as.matrix(questions[62:66,10:14]); colnames(q18) <- 1:5; q18
colnames(results2)[19]
q19 <- as.matrix(questions[67:71,10:14]); colnames(q19) <- 1:5; q19
colnames(results2)[20]
q20 <- as.matrix(questions[72:76,10:14]); colnames(q20) <- 1:5; q20
colnames(results2)[21]
q21 <- as.matrix(questions[77:81,10:14]); colnames(q21) <- 1:5; q21
colnames(results2)[22]
q22 <- as.matrix(questions[82:86,10:14]); colnames(q22) <- 1:5; q22
colnames(results2)[23]
q23 <- as.matrix(questions[92:96,10:14]); colnames(q23) <- 1:5; q23
colnames(results2)[24]
q24 <- as.matrix(questions[97:101,10:14]); colnames(q24) <- 1:5; q24
colnames(results2)[25]
q25 <- as.matrix(questions[102:106,10:14]); colnames(q25) <- 1:5; q25
colnames(results2)[26]
q26 <- as.matrix(questions[107:111,10:14]); colnames(q26) <- 1:5; q26
colnames(results2)[27]
q27 <- as.matrix(questions[112:116,10:14]); colnames(q27) <- 1:5; q27
colnames(results2)[28]
q28 <- as.matrix(questions[117:121,10:14]); colnames(q28) <- 1:5; q28
colnames(results2)[29]
q29 <- as.matrix(questions[122:126,10:14]); colnames(q29) <- 1:5; q29
colnames(results2)[30]
q30 <- as.matrix(questions[122:126,10:14]); colnames(q30) <- 1:5; q30
colnames(results2)[31]
q31 <- as.matrix(questions[127:131,10:14]); colnames(q31) <- 1:5; q31
colnames(results2)[32]
q32 <- as.matrix(questions[132:136,10:14]); colnames(q32) <- 1:5; q32
colnames(results2)[33]
q33 <- as.matrix(questions[137:141,10:14]); colnames(q33) <- 1:5; q33
colnames(results2)[34]
q34 <- as.matrix(questions[142:146,10:14]); colnames(q34) <- 1:5; q34
colnames(results2)[35]
q35 <- as.matrix(questions[147:151,10:14]); colnames(q35) <- 1:5; q35
colnames(results2)[36]
q36 <- as.matrix(questions[152:156,10:14]); colnames(q36) <- 1:5; q36
colnames(results2)[37]
q37 <- as.matrix(questions[157:161,10:14]); colnames(q37) <- 1:5; q37
colnames(results2)[38]
q38 <- as.matrix(questions[157:161,10:14]); colnames(q38) <- 1:5; q38
colnames(results2)[39]
q39 <- as.matrix(questions[162:166,10:14]); colnames(q39) <- 1:5; q39
colnames(results2)[40]
q40 <- as.matrix(questions[162:166,10:14]); colnames(q40) <- 1:5; q40
colnames(results2)[41]
q41 <- as.matrix(questions[167:171,10:14]); colnames(q41) <- 1:5; q41
colnames(results2)[42]
q42 <- as.matrix(questions[167:171,10:14]); colnames(q42) <- 1:5; q42
colnames(results2)[43]
q43 <- as.matrix(questions[172:176,10:14]); colnames(q43) <- 1:5; q43
colnames(results2)[44]
q44 <- as.matrix(questions[172:176,10:14]); colnames(q44) <- 1:5; q44
colnames(results2)[45]
q45 <- as.matrix(questions[177:181,10:14]); colnames(q45) <- 1:5; q45
colnames(results2)[46]
q46 <- as.matrix(questions[182:186,10:14]); colnames(q46) <- 1:5; q46
colnames(results2)[47]
q47 <- as.matrix(questions[187:191,10:14]); colnames(q47) <- 1:5; q47
colnames(results2)[48]
q48 <- as.matrix(questions[197:201,10:14]); colnames(q48) <- 1:5; q48
colnames(results2)[49]
q49 <- as.matrix(questions[202:206,10:14]); colnames(q49) <- 1:5; q49
colnames(results2)[50]
q50 <- as.matrix(questions[207:211,10:14]); colnames(q50) <- 1:5; q50
colnames(results2)[51]
q51 <- as.matrix(questions[207:211,10:14]); colnames(q51) <- 1:5; q51
colnames(results2)[52]
q52 <- as.matrix(questions[212:216,10:14]); colnames(q52) <- 1:5; q52
colnames(results2)[53]
q53 <- as.matrix(questions[217:221,10:14]); colnames(q53) <- 1:5; q53
colnames(results2)[54]
q54 <- as.matrix(questions[222:226,10:14]); colnames(q54) <- 1:5; q54
colnames(results2)[55]
q55 <- as.matrix(questions[227:231,10:14]); colnames(q55) <- 1:5; q55
colnames(results2)[56]
q56 <- as.matrix(questions[232:236,10:14]); colnames(q56) <- 1:5; q56
colnames(results2)[57]
q57 <- as.matrix(questions[242:246,10:14]); colnames(q57) <- 1:5; q57
colnames(results2)[58]
q58 <- as.matrix(questions[247:251,10:14]); colnames(q58) <- 1:5; q58
colnames(results2)[59]
q59 <- as.matrix(questions[252:256,10:14]); colnames(q59) <- 1:5; q59
colnames(results2)[60]
q60 <- as.matrix(questions[252:256,10:14]); colnames(q60) <- 1:5; q60
colnames(results2)[61]
q61 <- as.matrix(questions[87:91,10:14]); colnames(q61) <- 1:5; q61
colnames(results2)[62]
q62 <- as.matrix(questions[192:196,10:14]); colnames(q62) <- 1:5; q62
colnames(results2)[63]
q63 <- as.matrix(questions[212:216,10:14]); colnames(q63) <- 1:5; q63
colnames(results2)[64]
q64 <- as.matrix(questions[237:241,10:14]); colnames(q64) <- 1:5; q64
```

## Answers Data Frame Creation

```{r answers creation}
types <- data.frame("1.0"=rep(0,64),"0.8"=rep(0,64),"0.6"=rep(0,64),"0.4"=rep(0,64),"0.2"=rep(0,64))
colnames(types) <- c("1.0","0.8","0.6","0.4","0.2")
head(types)
```

## Test Answers

```{r testing}
m <- as.numeric(results[2,-c(1:3)])

for (i in 1:64) {
  temp <- get(paste("q",i,sep=""))
  types[i,] <- temp[,colnames(temp) == m[i]]
}
neuro <- data.frame("player"=rep(0,1),"type1a"=rep(0,1),"type1b"=rep(0,1),"type2a"=rep(0,1),
                    "type2b"=rep(0,1),"type3"=rep(0,1))
nt <- c("1a","1b","2a","2b","3")
for (i in 1:5) {
  neuro[1,i+1] <- sum(length(which(types$`1.0` == nt[i]))*1,length(which(types$`0.8` == nt[i]))*0.8,
                      length(which(types$`0.6` == nt[i]))*0.6,length(which(types$`0.4` == nt[i]))*0.4,
                      length(which(types$`0.2` == nt[i]))*.2)
}
neuro
```

## Create Function

```{r function}
player <- results[1,]
neuro[1,1] <- player[2]

neuro_types <- function(a) {
  types <- data.frame("1.0"=rep(0,64),"0.8"=rep(0,64),"0.6"=rep(0,64),"0.4"=rep(0,64),"0.2"=rep(0,64))
  colnames(types) <- c("1.0","0.8","0.6","0.4","0.2")
  total <- nrow(a)
  neuro <- data.frame("player"=rep(0,total),"type1a"=rep(0,total),"type1b"=rep(0,total),
                      "type2a"=rep(0,total),"type2b"=rep(0,total),"type3"=rep(0,total))
  for (i in 1:total) {
    athlete <- a[i,]
    neuro[i,1] <- athlete[2]
    player <- as.numeric(athlete[,-c(1:3)])
    for (j in 1:64) {
      temp <- get(paste("q",j,sep=""))
      types[j,] <- temp[,colnames(temp) == player[j]]
    }
    nt <- c("1a","1b","2a","2b","3")
    for (k in 1:5) {
      neuro[i,k+1] <- sum(length(which(types$`1.0` == nt[k]))*1,length(which(types$`0.8` == nt[k]))*0.8,
                          length(which(types$`0.6` == nt[k]))*0.6,length(which(types$`0.4` == nt[k]))*0.4,
                          length(which(types$`0.2` == nt[k]))*0.2)
    }
  }
  return(neuro)
}
```

## Function Test and Run with Team Results

```{r run with test data}
suppressWarnings(neuro_types(results)[,2:6])
```

Here we see that the largest result of the row is the primary neurotype of our athlete, followed by the second largest, ans so on.

```{r run with team data, include=FALSE}
NeuroTypes <- suppressWarnings(neuro_types(soccer_results))
soccer <- NeuroTypes[-c(1:4,23,28,29),]
write.csv(soccer, "neurotypes_soccer.csv", row.names = F)
save.image("~/MSBA/UT AD Work/Neurotyping/neurotyping function.RData")
```



